---
title: "Global Food Price Prediction"
author: "Jaimie Cairns, Dustin Cole, Morgan Hager-Perry"
date: "2025-03-26"
output:
  tufte::tufte_html:
    tufte_features: ["fonts", "italics"]
    df_print: paged
    toc: true # table of content true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false
  tufte::tufte_handout: default
  html_document:
    df_print: paged
    toc: true 
    toc_depth: 3  
    number_sections: true  
    theme: united  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tufte)
library(feasts)
library(tidyverse)
library(fpp3)
library(urca)
library(fracdiff)
library(glue)
library(GGally)
library(nycflights13)
library(dplyr)

library(quantmod)
library(zoo) 
library(tsibble)
```

# Food Prices Dataset

## Time series decomposition

### Import Data

```{r}

#Data from Bureau of Labor Statistics:  Food and beverages in U.S. city average, all urban consumers, not seasonally adjusted. Downloaded all monthly data from 1967 forward from https://data.bls.gov/dataViewer/view/timeseries/CUUR0000SAF;jsessionid=40005007FB9F133245ECFC6162596EB0

#input the data, mutate the time column to yearmonth and convert to a tsibble
food_prices <- readr::read_csv("file.csv", show_col_types=FALSE) |>
  mutate(Month=yearmonth(Label))|> 
  filter(year(Month)>1999)|>
  select(Month, Value) |>
  as_tsibble(index=Month)

food_prices<-na.locf(food_prices)
#View tsibble
print(food_prices)
```

### Classic Addditive Decomposition

```{r}
food_prices |>
  model(
    classical_decomposition(Value, type = "additive")
  ) |>
  components() |>
  autoplot() +
  labs(title = "Classical additive decomposition of US Food & Beverage Prices")
```

## Time series Visualization

### Visualizations: Not adjusted for inflation 

#### Time Plot

```{r}
food_prices|>
  autoplot(Value) + 
  labs( y="US Dollar", title="US Food Prices")
```

#### Autocorrelation Function

```{r}
food_prices|>
  ACF(Value)|>
  autoplot() +labs(title="US Food Prices ACF")
```

#### Seasonal Plot
```{r}
food_prices|>
  gg_season(Value, labels="both")+
  labs(y="US $", title="Seasonal plot: US Food and Beverage Prices")
```

#### Seasonal subseries plot
```{r}
food_prices |>
  gg_subseries(Value)+
  labs(y = "US$", title="US Food and Beverage Costs")
```


### Visualizations: Inflation Adjusted


```{r}

#import CPI data for US economy from BLS: 
#https://fred.stlouisfed.org/series/CPIAUCNS

us_economy <- readr::read_csv("CPIAUCNS.csv", show_col_types=FALSE) |>
  mutate(Month=yearmonth(observation_date))|> 
  filter(year(Month)>2000)|>
  select(CPIAUCNS, Month) |>
  as_tsibble(index=Month)

print(us_economy)
```
### Time Plot Inflation Adjusted vs. Non-Adjusted

```{r}
food_prices|>
  left_join(us_economy, by="Month") |>
  mutate(Adjusted_Value= Value/CPIAUCNS *100) |>
  pivot_longer(c(Value, Adjusted_Value), values_to="Value")|>
  mutate(name=factor(name, 
                     levels=c("Value","Adjusted_Value"))) |>
  ggplot(aes(x=Month, y=Value))+
  geom_line()+
  facet_grid(name ~.,scales="free_y")+
  labs(title="US Food & Beverage Prices",y="US$")
```
#### Seasonal Plot Inflation Adjusted

```{r}
#created adjusted food price tsibble
food_prices_adj<-food_prices|>
  left_join(us_economy, by="Month") |>
  mutate(Adjusted_Value= Value/CPIAUCNS *100) |>
  select(Adjusted_Value,Month)

#remove nas
food_prices_adj<-na.locf(food_prices_adj)

#seasonal plot of adjusted food prices

food_prices_adj|>
  gg_season(Adjusted_Value, labels="both")+
  labs(y="US $", title="Seasonal plot: US Food and Beverage Prices")
```
#### Autocorrelation Function Inflation Adjusted

```{r}
food_prices_adj|>
  ACF(Adjusted_Value)|>
  autoplot() +labs(title="US Food Prices ACF")
```

#### Seasonal subseries plot
```{r}
food_prices_adj |>
  gg_subseries(Adjusted_Value)+
  labs(y = "US$", title="US Food and Beverage Costs")
```

### Decomposition of Inflation Adjusted Data

```{r}
food_prices_adj |>
  model(
    classical_decomposition(Adjusted_Value, type = "additive")
  ) |>
  components() |>
  autoplot() +
  labs(title = "Classical additive decomposition of Inflation Adjusted US Food & Beverage Prices")
```

## Description of Food Prices time series

The US Food and Beverage dataset shows a clear upward trend in the cost of food and beverages from 2000 to present. In looking at the unadjusted data the trend appears significant and constant; however, when data is adjusted for inflation the trend is still upward but appears to have more cyclical movement.

While the decomposition graphs of both the unadjusted and adjusted data show what appears to be significant seasonality in the data, the seasonality is somewhat minor. When the seasonal subseries plot is looked at in the unadjusted data, there appears to be almost no variation in monthly costs even though you would expect to see that with seasonally impacted data. The seasonality is a bit more noticeable when the seasonal subseries plot is shown for the adjusted for inflation data. Here it becomes apparent that food costs are lower in the summer months and higher in the winter months.


# Stock Index Dataset

## Import and set up
```{r}
#pulling the stock ticker for SP500 Food Products Index - Same time range as other dataset
getSymbols("^SP500-302020",
           src="yahoo",
           from = "2000-01-07",
           to = "2025-01-31")


# Get the closing prices as an xts object
stock_data <- Cl(`SP500-302020`)  

# Create a complete time sequence including weekends
all_dates <- seq.Date(from = as.Date("2000-01-07"), to = as.Date("2025-01-31"), by = "day")
stock_data_full <- merge(stock_data, xts(, all_dates))  # Merge with full date sequence

#adjusting the column names 
colnames(stock_data_full) <- c("Close")

# Check the result
head(stock_data_full)
```
### Fill missing values
```{r}
# Forward fill the missing values
stock_data_filled <- na.locf(stock_data_full)

# View the data after filling missing values
head(stock_data_filled)
```
### plotting the time series
```{r}
plot(stock_data_filled)
```
## Additive Decomposition Plot
```{r}
SP500_ts <- ts(stock_data_filled,frequency=365)
plot(decompose(SP500_ts))
```

## Visualizations
### Visualization 1 - Trend
```{r}
SP500_ts.decom <- decompose(SP500_ts, type = "mult")
plot(SP500_ts.decom)
Trend <- SP500_ts.decom$trend
Seasonal <- SP500_ts.decom$seasonal
ts.plot(cbind(Trend, Trend * Seasonal), lty=1:2)
```

### Visualization 2 - ACF Plot
```{r}
acf(SP500_ts)
```

### Visualization 3 - Seasonality
```{r}
# Convert xts to tsibble
df <- data.frame(Date = zoo::index(stock_data_filled), Close = zoo::coredata(stock_data_filled))
tsibble_data <- as_tsibble(df, index = Date)

#plot the seasonality chart
tsibble_data %>%
  gg_season(Close, labels = "both") +
  labs(y = "Closing Price ($)",
       title = "Seasonal plot: S&P500 Food Price Index")
```

### Visualization 4 - Moving Averages
```{r}
#take a subset of the data for readability of chart
tsibble_data_subset <- tail(tsibble_data,100)

#create moving averages
tsibble_data_ma <- tsibble_data_subset |>
  mutate(
    `5-MA` = slider::slide_dbl(Close, mean,
                               .before = 2, .after = 2, .complete = TRUE),
    '7-MA' = slider::slide_dbl(Close, mean,
                               .before = 3, .after = 3, .complete = TRUE)            
    
  )

#plot the data
tsibble_data_ma |>
  autoplot(Close) +
  geom_line(aes(y = `5-MA`), colour = "orange") +
  geom_line(aes(y = `7-MA`), colour = "green") +
  labs(y = "Closing Price ($)",
       x = "Month",
       title = "Closing Price for S&P500 Food Price Index")
```

## Description of Stock Index Time Series:

As we would expect from data on a stock index, there is a clear upward trend over the last 25 years. This would be seen by almost all sectors of the S&P500, where there is flucutation within date ranges, but over a long time horizon (i.e. 20+ years) you will almost certainly see an increasing trend. Surprisingly, the decomposition showed that there is some seasonality to the stock index which isn't as pronounced when looking at the seasonality chart (visualization 3). I can see some seasonality in that there are a few years where there is downturns around July and most years increase the most in December. This dataset is certainly non-stationary as there is a clear increasing trend therefore the mean will change over time (as seen in the moving average chart - visualization 4).
